Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-16 treestats_paper.tex
fixed error in branch defn
clarifying units of duality

Equation~\eqref{eqn:branch_stat_tree} defines a statistic for a single tree,
but in practice it is more useful to average branch statistics
over a region of the genome. We do this by averaging the tree statistics over the region
with probabilities proportional to the trees' spans:
\begin{align}
    \branch(f, \iw)_{[c,d)}
    &=
    \frac{1}{d-c} \sum_{k=1}^{n(\treeseq)} \ell_k(c,d) \branch(f, \iw)_{T_k} ,
\end{align}
where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ covers
(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
The \emph{polarised} version of a Branch statistic
is defined analogously to the Node and Site versions:
\begin{align} \label{eqn:branch_polarised}
    \text{(polarised)} \qquad
    \branchp(f, \iw)_T
    &=
    \sum_{u \in T} f(\nw_{T}(u)) ,
\end{align}
and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
However, the correspondence between the polarised statistics is less tight
than for the unpolarised ones.

@@

This relationship is true more generally.
In fact, for any region of the genome between $a$ and $b$,
with $\treeseq_{[a,b)}$ the trees spanning that interval,
\begin{align}
    \branch(f, \iw)_{[a,b)}
    =
    \E\left[ \site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)} \right] ,
\end{align}
where the expected value averages over infinite-sites mutations with rate 1
(and so lengths are measured in units of expected numbers of mutations)
but leaves the genealogies fixed.
In other words, the Site statistic can be thought of an estimator of its corresponding Branch statistic,
which is itself a summary of local tree shapes.
Let's unpack the assumptions here: what exactly is the mutational model?
First, we are taking the mutation rate to be 1, i.e.,
the expected number of mutations that occur on a region of the genome of length $\ell$
over $t$ units of time is equal to $\ell t$.
This just amounts to a change of units ---
for instance, if genome length is measured in nucleotides,
and the probability of mutation is $\mu$ per nucleotide and per generation,
then we are measuring times in units of $1/\mu$ generations.
Second, we are assuming that the probability of per-site mutation is low enough
that no two mutations occur at the same site
--- the fact that they do, occasionally, means that this is an approximation.
Third, we are assuming that mutation rates are constant through time and across the genome.
Of course, the statement remains true if we can measure distance along the genome and time
in a way that mutation rates are constant, but how these vary is generally unknown.

>>>>

Equation~\eqref{eqn:branch_stat_tree} defines a statistic for a single tree,
but in practice it is more useful to average branch statistics
over a region of the genome. We do this by averaging the tree statistics over the region
with probabilities proportional to the trees' spans:
\begin{align}
    \branch(f, \iw)_{[c,d)}
    &=
    \frac{1}{d-c} \sum_{k=1}^{n(\treeseq)} \ell_k(c,d) \branch(f, \iw)_{T_k} ,
\end{align}
where $\ell_k(c,d)$ is the length of the region in $[c,d)$ that the tree $T_k$ covers
(i.e., if $T$ covers the half-open interval $[a_{k-1},a_k)$,
then $\ell_k(c,d) = \max(0, \min(d,a_k) - \max(c,a_{k-1}))$).
The \emph{polarised} version of a Branch statistic
is defined analogously to the Node and Site versions:
\begin{align} \label{eqn:branch_polarised}
    \text{(polarised)} \qquad
    \branchp(f, \iw)_T
    &=
    \sum_{u \in T} \beta_T(u) f(\nw_{T}(u)) ,
\end{align}
and $\branchp(f, \iw)_{[c,d)}$ is defined using $\branchp(f, \iw)_T$ as before.
However, the correspondence between the polarised statistics is less tight
than for the unpolarised ones.

@@

This relationship is true more generally.
In fact, for any region of the genome between $a$ and $b$,
with $\treeseq_{[a,b)}$ the trees spanning that interval,
\begin{align}
    \branch(f, \iw)_{[a,b)}
    =
    \frac{1}{\mu}
    \E\left[ \site(f, \iw)_{[a,b)} \given \treeseq_{[a,b)} \right] ,
\end{align}
where the expected value leaves the genealogies fixed,
averaging over infinite-sites mutations
with rate $\mu$ per unit time and per unit of sequence length.
In other words, the Site statistic can be thought of an estimator of its corresponding Branch statistic,
which is itself a summary of local tree shapes.
Because we have normalised these statistics by the length of the genome under consideration,
both sides of this equation are in units of \emph{time}:
Branch statistics give mean weighted branch lengths;
Site statistics give mean densities of mutations per unit of sequence length,
which divided by $\mu$ is converted to time.
Let's unpack the assumptions here: what exactly is the mutational model?
First, we are taking the mutation rate to be $\mu$, i.e.,
the expected number of mutations that occur on a region of the genome of length $\ell$
over $t$ units of time is equal to $\mu \ell t$.
Second, we are assuming that the probability of per-site mutation is low enough
that no two mutations occur at the same site
--- the fact that they do, occasionally, means that this is an approximation.
Third, we are assuming that mutation rates are constant through time and across the genome.
Of course, the statement remains true if we can measure distance along the genome and time
in a way that mutation rates are constant, but how these vary is generally unknown.
